---
title: "Lab Exercises 1 Solutions"
output:
  html_document: default
  html_notebook: default
---

## Question 1

Andrew Johns has played 153 games for the Newcastle Knights NRL team; of those 100 games have been wins for the Knights. During that time, Newcastle have played 223 games in total, winning 135.

#### (a) Estimate the following:

##### (i) Pr(Johns plays)
153/223 = `r 153/223`

##### (ii) Pr(Knights win)
135/223 = `r 135/223`

##### (iii) Pr(Johns plays AND Knights win)
Johns has played in 100 games where the Knights have won, so answer 100/223 = `r 100/223`

##### (iv) Pr(Johns played | Knights win)
We just read from the table that = 100/135 OR using definition
Pr (Johns played | Knights win) = Pr (Johns played, Knights win)/ Pr (Knights win) =	(100/223)/(135/223) = 100/135 = 0.448/0.605 = `r 0.448/0.605`

##### (v) Pr(Knights win | Johns played)
Pr (Knights win | Johns played) = 100/153 = `r 100/153`
Or using definition = Pr (Johns played, Knights win)/ Pr (Johns played) = 100/223/(153/223)

##### (vi) Pr(Knights win | Johns didn't play)
Pr(Knights win | Johns didn't play) = 35/70 = 0.5 OR using definition
=	Pr(Knights win, Johns didn't play)/Pr(Johns didn't play) = 35/223/(70/223) = 0.5

##### (vii) Pr(Johns played | Knights lose)	
Pr(Johns played | Knights lose) = Pr (Johns played, Knights lose)/ Pr (Knights lose)

Firstly, Johns has played in 153-100=53 games the Knights have lost, so Pr (Johns played, Knights lose) = 53/223 = 0.238.
Pr(Knights lose) = (223-135)/223 = 88/223 = 0.395

So, Pr(Johns played | Knights lose) = 53/88 = 0.603 (this can be read simply from the table)
						
Johns played	Johns didn't play	Total		
Knights win	100	35	135		
Knights lose	53	35	88		
Sum		153	70	223		

#### (b)	Think about the following:

##### 1)	Interpret the probabilities in (iii), (iv), (v), (vi) and (vii).
Firstly, of the 223 games the Knights have played, the events that Johns played and Knights won both occurred 44.8% = 100*(100/223) of the time. For games in which we know the Knights won, Johns played in about 74% of them; for games in which we know Johns played, the Knights won 65%; while if Johns didn't play the Knights won only 35% of these. Finally for games the Knights lost, Johns played in 60% of them.

##### 2)	How could each be used in footy tipping or betting?
Only the 65% = Pr (Knights win | Johns played) and the 50% Knights win when Johns doesn't play should be used in betting. Betting is equivalent to forecasting. We can not know the result (Knights win or lose) before the game, we can however know that Johns is listed to play in an upcoming match. If that is the case, the Knights have won 65% of those matches, compared to only 60% overall and 50% wins when Johns did not play. We are more likely to bet on the Knights if Johns is playing.

##### 3)	Do we need to know 223 (the total number of games) to calculate these?
No: 223 cancels out in all calculations except in (iii).
 
## Question 2 
Consider the following example, adapted from Albert (1997) Bayesian computation using Minitab:

A common blood test for AIDS has a false negative rate of 3% (i.e. if you have the disease, the test comes up negative 3% of the time) and a false positive rate of 2%. AIDS has an incidence 1 in 200 in Australia. You take a blood test and it comes up positive, what is the probability that you actually have AIDS? (The questions below will guide you through the answer.)

#### (a) Discussion: What are the relevant events A and B here? What prior information do we have? What is Pr(A)?
Bayes' rule updates Pr(A) given the info in B to give Pr(A \mid B). The events are: 

A: you have AIDS.
B: you return a positive blood test.

We know that $Pr(Bc \mid A) = 0.03$ (false negative) AND $Pr(B \mid Ac) = 0.02$ (false positive). We also need to know Pr(A). This is the prior probability that you have the disease, before the blood test. This could be taken as the population incidence of the disease OR some estimate based on your family history, risk factors etc.

Let's estimate	$Pr(A) = 1/200 =$ `r 1/200`

#### (b) Use Bayes' rule to update the prior probability, Pr(A).

$$
  \begin{aligned}
    Pr(A \mid B) 
    &=	\frac
          {Pr(B \mid A) Pr(A)}
          {Pr(B)} \\ 
    & =	\frac
          {Pr(B \mid A) Pr(A)}
          {Pr(B \mid A) Pr(A) + Pr(B \mid A^{c}) Pr(A^{c})} \\ 	
    &=	\frac
          {1}
          {1 + \frac
                  {Pr(B \mid A) Pr(A^{c})Pr(A^{c})}
                  {Pr(B\mid A) Pr( A)}}
  \end{aligned}
$$

Notice that the denominator term $Pr(B)$ is NOT needed (but can be deduced if necessary) in this calculation. $Pr(A \mid B) = \frac{1}{1 + \frac{0.02 \times 0.995}{(1 - 0.03) \times 0.005}}	= 0.196$			

Clearly you should NOT be too worried about a positive result, you still have an 80% chance it is a false positive.

#### (c) Prepare a table showing the prior, likelihood and posterior probabilities of each event.

Model	Prior	Likelihood	Product	Posterior
		of +ve test		
Have disease	0.005	0.97	0.00485	0.196
Dont	0.995	0.02	0.0199	0.804
SUM			0.02475	1

In R: 
```{r}
prior=c(0.005,0.995) 
likelihood=c(0.97,0.02) 
prod=prior*likelihood 
sum=sum(prod) 
posterior=prod/sum
```

#### (d) How strongly did the data affect the prior in this case?
The prior probability of having the disease was 0.005. The posterior is 0.196. This is an increase of .196/.005 = 39 times. One positive blood test increases your chances of having the disease by a factor of 39. However, it is still not a large number.

#### (e) How high would Pr(A) have to be for you to be worried about a positive result?

Let's examine three other priors.

##### (i)
Say the incidence of AIDS was 1 in 100

Model	Prior	Likelihood	Product	Posterior
		of +ve test		
Have disease	0.01	0.97	0.0097	0.329
Don't	0.99	0.02	0.0198	0.671
SUM			0.0295	1

In R: 

```{r}
prior=c(0.01,0.99) 
likelihood=c(0.97,0.02) 
prod=prior*likelihood 
sum=sum(prod) 
posterior=prod/sum
```

##### (ii) What about incidence 1 in 20?

Model	Prior	Likelihood	Product	Posterior
		of +ve test		
Have disease	0.05	0.97	0.0485	0.719
Don't	0.95	0.02	0.019	0.281
SUM			0.0675	1


In R: 

```{r}
prior=c(0.05,0.95) 
likelihood=c(0.97,0.02) 
prod=prior*likelihood 
sum=sum(prod) 
posterior=prod/sum
```

##### (iii)	Your partner has this disease and a doctor estimates that Pr(A) = 0.2

Model	Prior	Likelihood	Product	Posterior
		of +ve test		
Have disease	0.2	0.97	0.194	0.924
Don't	0.8	0.02	0.016	0.076
SUM			0.21	1


In R: 

```{r}
prior=c(0.2,0.8)
likelihood=c(0.97,0.02)
prod=prior*likelihood
sum=sum(prod)
posterior=prod/sum
```

Once the incidence or prior probability of the disease gets above about 1 in 100, you start to get worried.

#### (f) Suppose you have 3 separate positive blood test results. What is your updated probability of having the disease?

Using the same table idea and assuming independence of tests (is that a valid assumption?),

```{r}
names <- c("Model",	"Prior", "Likelihood of \n 3 +ve tests",	"Product",	"Posterior")
have <- c(0.005,0.973,0.004563365,0.998)
dont<- c(0.995,0.023,0.00000796,0.00174)
sum <- c("","",0.004571325,1)
data <- rbind(have,dont,sum)
rnames <- c("Have Disease","Don't","Sum")
data <- cbind(rnames,data)
colnames(data) <- c(names)
```

Clearly, repeated positive tests add lots of evidence in favour of you having the disease.

In R: 
```{r}
prior=c(0.005,0.995)
likelihood=c((0.97)^3,(0.02)^3)
prod=prior*likelihood
sum=sum(prod)
posterior=prod/sum
```

## Question 3
Three approaches: experts, empirical analysts and logicians.

BDA, exercise 1.10, no.7. At the end of a game show called Let's make a deal, contestants are asked to choose 1 of 3 doors. Behind 1 door is a car, behind the other 2 doors are lesser prizes. After the contestant chooses a door, Monte Hall (the host) opens a door that he knows has a lesser prize behind it (NOT the door chosen by the contestant). Monte then offers the contestant to switch to the 3rd remaining door OR stick with the original choice of door. Should the contestant switch doors?

#### Expert (mathematical) approach:
There are three variables here. Let C = the door number that the car is behind (1,2,3); Y = the door number that you choose; and let M = the door number that the compere decides to open. Without loss of generality assume that you choose door 1 first. What is $Pr(C=1\mid M=2, Y=1)$?

There are three variables here. Let
C = the door number that the car is behind (1,2,3);
Y = the door number that you choose; and
M = the door number that the compere decides to open.

Without loss of generality assume that you choose door 1 first. Monte Hall can now choose either door 2 or 3, depending where the car is. Let's say he chooses door 2 (i.e. the car can only be behind door 1 or 3).
$$
  \begin{aligned}
    Pr(C = 1\mid M = 2, Y = 1) 
      &= \frac
          {Pr(C = 1,M = 2, Y = 1)}
          {Pr(M = 2, Y = 1)} \\
      &= \frac
            {Pr(M = 2\mid C = 1, Y = 1)Pr(C = 1\mid Y = 1)P(Y = 1)}
            {Pr(M = 2\mid Y = 1)P(Y = 1)} \\
      &=  \frac
            {Pr(M = 2\mid C = 1, Y = 1)Pr(C = 1\mid Y = 1)}
            {Pr(M = 2 \mid Y = 1)} \\
      &=  \frac
            {Pr(M = 2\mid C = 1, Y = 1)Pr(C = 1\mid Y = 1)}
            {Pr(M = 2\mid C = 1, Y = 1)Pr(C = 1\mid Y = 1) + Pr(M = 2\mid C = 2, Y = 1)Pr(C = 2\mid Y = 1) + Pr(M = 2\mid C = 3, Y = 1)Pr(C = 3\mid Y = 1)} \\
      &=  \frac
            {1/2 \times 1/3}
            {1/2 \times 1/3 + Pr(M = 2\mid C = 2, Y = 1)Pr(C = 2\mid Y = 1) + Pr(M = 2\mid C = 3, Y = 1)Pr(C = 3\mid Y = 1)} \\
      &=	\frac
            {1/6}
            {1/6 + 0 \times 1/3 + 1 \times 1/3} \\
      &= \frac
            {1}
            {3}	
	\end{aligned}
$$

Note that $Pr(M = 2 \mid C = 3, Y = 1) = 1$ as Monte Hall must choose door 2 in this case.

We know C can't equal 2 as the compere cannot choose the door the car is behind, i.e. $Pr(C = 2 \mid M = 2, Y = 1) = 0$.

The other case is that C=3, with a similar way to the above such that,

$$
  \begin{aligned}
    Pr(C = 3 \mid M	= 2, Y = 1)	
      &= \frac
            {Pr(M = 2\mid C = 3, Y = 1)Pr(C = 3\mid Y = 1)}	
					  {Pr(M = 2\mid Y = 1)} \\
			&=  \frac
			      {1 \times 1/3}
			      {1/6 + 1/3} \\
			&=	2
	\end{aligned}
$$

Clearly the contestant should always swap doors when given the chance.
 
#### Logical approach:

What is the probability of the chosen door having the car behind it? What is the probability of the door that the compere opens having the car behind it? Does the probability for the chosen door change when the compere opens a door? Combine this information to decide if the contestant should change doors.

When the contestant chooses a door, it is 1/3 likely to have the car behind it. This does NOT change when Monte Hall opens another door, since Monte knows where the car is and always chooses a door without the car (this door has probability 0 of having the car behind it). Thus the door not chosen by the contestant (OR Monte Hall) has probability 2/3 of containing the car. The contestant should always SWAP doors, giving twice the probability of winning the car.

#### Empirical approach:

Do the experiment about 10--20 times with 3 cups and a coin or ping-pong ball, or use the utility 'Let's make a deal' (e.g. http://math.ucsd.edu/\simcrypto/Monty/monty.html).

You will see you win twice as often if you switch cups.

End of Lab 1