---
title: "Assignment 3 - Question 2"
output:
  html_document: default
  html_notebook: default
---
```{r setup, include=FALSE}
require(tufte)
require(knitr)
opts_chunk$set(cache = FALSE)
opts_chunk$set(fig.align = 'center', fig.height = 3.5, fig.width = 3.5, collapse = TRUE, highlight = TRUE)
```
```{r data, include=FALSE}
data <- readRDS("C:/R/STAT3120/Assignments/Assignment3/Data/birthweight.rds")
```

##Question 2

**Taken from Albert (2007). Normal linear regression.**

Dobson (2001) describes a birthweight regression study. One is interested in predicting a baby’s birthweight (in grams) based on the gestational age (in weeks) and the gender of the baby. The data is presented in the file *birthweight.txt*. In the standard linear regression model, we assume that
$$
\text{Birthweight (i)} = \beta_{0} + \beta_{1}\,AGE\,(i) + \beta_{2}\,GENDER\,(i) + e(i)
$$
where the *e(i)’s* are independent and Normally distributed random variables with mean **0** and variance $\sigma^{2}$.

#####(a) Use the R function lm to fit this model by Least-Squares. From the output, assess if the effects AGE and GENDER are significant, and if they are significant, describe the effects of each covariate on birthweight.

#####(b)	Suppose a uniform prior is placed on the regression parameters $\beta$ = ($\beta_{0}$, $\beta_{1}$, $\beta_{2}$). Use the LearnBayes package in R (blinreg command) to simulate an MC sample of **5000** draws from the joint posterior distribution of ($\beta$, $\sigma^{2}$). From the simulated sample, compute posterior means and standard deviations of $\beta_{1}$ and $\beta_{2}$. Compare the results with the LS estimates obtained in **(a)**.

#####(c) Repeat part **(b)** using WinBUGS. Compare the results with the LS estimates obtained in **(a)** and MC sampling in part **(b)**.
